
\documentclass[conference]{IEEEtran}



\usepackage{subfigure}
\usepackage{graphicx}
\usepackage{algorithm,algorithmic,epsfig}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{verbatim}
\usepackage{array}
\usepackage[hmargin=0.585in,vmargin=0.95in]{geometry}
\usepackage{comment}
\usepackage{soul}
\usepackage{color}
%\usepackage{multirow}
\usepackage{balance}
%\usepackage{picins}
\makeatother

%\renewcommand{\figurename}{Figure}

\begin{document}

\title{Advancing Intrusion Detection Systems: A Comprehensive Survey on AI, ML, and DL Approaches}

\author{\authorblockN{Vigneshwar Sundararajan and Nabhan Aziz}
\authorblockA{Department of Computer Science: Cybersecurity and Privacy\\
University of Central Florida\\
%Orlando, Florida\\
Email: \{vi126747, na734868\}@ucf.edu}}


\maketitle
\begin{abstract}

The rapid evolution of cyber threats has necessitated the development of sophisticated Intrusion Detection Systems (IDS) to safeguard critical infrastructures and digital assets. Artificial Intelligence (AI), Machine Learning (ML), and Deep Learning (DL) have emerged as transformative technologies, driving significant advancements in IDS capabilities. This survey paper provides a comprehensive review of state-of-the-art AI/ML/DL methodologies applied to intrusion detection, examining their effectiveness, limitations, and potential for addressing emerging security challenges. 

By analyzing 20 seminal research contributions, this study categorizes IDS approaches based on their underlying models, evaluation metrics, and datasets, offering insights into the comparative performance of techniques such as supervised learning, anomaly detection, and hybrid systems. Additionally, the paper highlights trends, including the integration of Explainable AI (XAI), the use of adversarial robustness mechanisms, and the adoption of real-time detection frameworks. Through critical analysis, this survey identifies research gaps, such as scalability, dataset diversity, and deployment challenges, while proposing future directions to enhance the efficacy and reliability of IDS. This work aims to serve as a valuable resource for researchers and practitioners in the pursuit of resilient cybersecurity solutions.

\textit{Index Terms} Intrusion Detection Systems, AI Security, Network Security, Adversarial Machine Learning, Real-time Detection, Hybrid Models.

\end{abstract}


\IEEEpeerreviewmaketitle

\input{Introduction}
\input{FoundationalConcepts}
\input{Methodology}
\input{LiteratureReview}
\input{Discussion}
\input{Conclusion}

%\IEEEtriggeratref

\bibliographystyle{IEEEtran}
\balance
\bibliography{TPMobBib}
%\input{biographies}


\end{document}


